---
title: "A framework to investigate trophic reorganization in marine systems using predator stomach contents"
subtitle: "Tables"
format:
  docx: default
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(readr); library(purrr)
})
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      results = "asis")

# flextable gives proper Word tables; knitr::kable is the fallback.
USE_FT <- requireNamespace("flextable", quietly = TRUE)

# Writes the table straight into the document. Explicit output (rather than
# relying on auto-printing) is what allows several tables in one chunk.
show_tbl <- function(d, digits = 1) {
  d <- d %>% mutate(across(where(is.double), ~ round(.x, digits)))
  if (USE_FT) {
    ft <- flextable::flextable(d)
    ft <- flextable::theme_booktabs(ft)
    ft <- flextable::fontsize(ft, size = 9, part = "all")
    ft <- flextable::font(ft, fontname = "Calibri", part = "all")
    # Span the page width and wrap text inside cells, so that wide tables
    # (long result sentences) never run off the page.
    ft <- flextable::set_table_properties(ft, layout = "autofit", width = 1)
    flextable::flextable_to_rmd(ft)
  } else {
    cat(knitr::kable(d), sep = "\n")
  }
  cat("\n\n")
  invisible(NULL)
}

# Reads a result file; returns NULL rather than stopping, so a table whose
# analysis has not been run shows a visible placeholder instead of breaking
# the whole document.
rd <- function(path) {
  if (file.exists(path)) read_csv(path, show_col_types = FALSE) else NULL
}
absent <- function(path) {
  cat("\n*[Missing: `", path, "` — run the corresponding analysis.]*\n\n", sep = "")
}

FAMILIES <- c("Stability", "Substitution", "Functional change", "Reorganization")
CONTRAST_LAB <- c(P1a = "2004 vs 2006", P1b = "2004–2005 vs 2006",
                  P2 = "2018 vs 2019", PTb = "2006 vs 2018",
                  PTa = "2004–2006 vs 2018–2019")
MODE_LAB  <- c(biomass = "Biomass", occurrence = "Occurrence")
LEVEL_LAB <- c(all_gulf = "Gulf-wide", ecoregion = "Ecoregion", stratum = "Stratum")

relabel <- function(d) {
  if ("family" %in% names(d))
    d <- d %>% mutate(family = recode(family, Reorganisation = "Reorganization"),
                      family = factor(family, levels = FAMILIES))
  if ("contrast" %in% names(d))
    d <- d %>% mutate(contrast = factor(contrast, levels = names(CONTRAST_LAB),
                                        labels = unname(CONTRAST_LAB)))
  if ("mode" %in% names(d))
    d <- d %>% mutate(mode = factor(mode, levels = names(MODE_LAB),
                                    labels = unname(MODE_LAB)))
  d
}

fmt <- function(x, d = 1) formatC(x, format = "f", digits = d)
```

# Main tables

**Table 1.** Diagnostic typology. Each predator × size-class cell is scored on three tests — prey composition (C, PERMANOVA), diversity (*H*′, jackknifed Shannon) and niche breadth (*B*s, standardized Levins) — as significant (sig, *p* < 0.05) or non-significant (ns). A composition trend (0.05 ≤ *p* < 0.10) is used only when *H*′ and *B*s are both non-significant; otherwise the cell is classified on the significant signal.

```{r t1}
tribble(
  ~`Diagnostic state`,            ~C,            ~`H′`, ~Bs,   ~Family,
  "Stable Diet",                  "ns (p ≥ 0.10)", "ns", "ns", "Stability",
  "Emerging Shift",               "trend",        "ns",  "ns",  "Stability",
  "Ghost Shift",                  "sig",          "ns",  "ns",  "Substitution",
  "Internal Rebalancing",         "ns",           "sig", "ns",  "Functional change",
  "Niche Compression/Expansion",  "ns",           "ns",  "sig", "Functional change",
  "Niche Restructuring",          "ns",           "sig", "sig", "Functional change",
  "Partial Diet Shift",           "sig",          "sig", "ns",  "Reorganization",
  "Structural Shift",             "sig",          "ns",  "sig", "Reorganization",
  "Major Shift",                  "sig",          "sig", "sig", "Reorganization"
) %>% show_tbl()
```

**Table 2.** Number of prey categories produced by each grouping architecture at reference aggregation thresholds *x* (minimum number of stomachs). Counts are distinct category labels across all predators. The per-predator architecture is swept over 5–250 stomachs only, so higher thresholds are not defined for it.

```{r t2}
dc_path <- "data/dat_classed.rda"
if (!file.exists(dc_path)) {
  absent(dc_path)
} else {
  e <- new.env(); load(dc_path, envir = e); dc <- get(ls(e)[1], envir = e)
  X_REF <- c(10, 250, 440, 1000)
  FAMS  <- c(`Pooled (_1)` = "_1", `Pooled cross-predator (_2)` = "_2",
             `Per-predator (_PP)` = "_PP")
  t2 <- map_dfr(names(FAMS), function(nm) {
    vals <- map_chr(X_REF, function(x) {
      col <- paste0("prey_category_", x, FAMS[[nm]])
      if (col %in% names(dc)) as.character(n_distinct(na.omit(dc[[col]]))) else "—"
    })
    tibble(`Grouping architecture` = nm, !!!setNames(as.list(vals), paste0("x = ", X_REF)))
  })
  show_tbl(t2)
}
```

# Supplementary tables — robustness of the diagnostic assignment

**Table S7.** Summary of the robustness checks. Each row states the objection addressed and the result at the Gulf-wide scale unless indicated. Values are read directly from the analysis outputs.

```{r ts7}
num <- function(d, ...) if (is.null(d)) NA_real_ else d %>% filter(...) %>% pull(1)

m3  <- rd("Robustness_all_gulf_1_T/M3_CI_overlap_verdict.csv")
m2  <- rd("Robustness_all_gulf_1_T/M2_FDR_family_frequencies.csv")
m4  <- rd("Robustness_all_gulf_1_T/M4_alpha_trend_sensitivity.csv")
m5  <- rd("Robustness_all_gulf_1_T/M5_stability_summary.csv")
nfp <- rd("NullCalibration_all_gulf_1/null_false_positive_rates.csv")
cvn <- rd("Calibrated_all_gulf_1/calibrated_vs_nominal.csv")
csv <- rd("Calibrated_all_gulf_1/calibrated_separation_verdict.csv")
bal <- rd("Robustness_all_gulf_1_BAL/balanced_separation_verdict.csv")
eff <- map_dfr(names(LEVEL_LAB), function(l) {
  d <- rd(sprintf("Effort_%s_1/C_separation_verdict.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})

res <- function(ok, txt) if (isTRUE(ok)) txt else "Not run"

r_boot <- res(!is.null(m3), sprintf(
  "Stability bands separated by %s (biomass) and %s (occurrence) points",
  fmt(m3$gap_pp[m3$mode == "biomass" & m3$family == "Stability"]),
  fmt(m3$gap_pp[m3$mode == "occurrence" & m3$family == "Stability"])))

r_fdr <- res(!is.null(m2), {
  s <- m2 %>% filter(contrast == "PTa", family == "Substitution", mode == "biomass")
  sprintf("Substitution %s%% → %s%% (biomass)", fmt(s$pct_raw), fmt(s$pct_bh))
})

r_alpha <- res(!is.null(m4), {
  g <- m4 %>% filter(trend == 0.10, alpha <= 0.075, family == "Stability") %>%
    group_by(mode, alpha, contrast_type) %>%
    summarise(lo = min(pct_mean), hi = max(pct_mean), .groups = "drop") %>%
    pivot_wider(names_from = contrast_type, values_from = c(lo, hi)) %>%
    mutate(gap = lo_within - hi_between)
  sprintf("Within and between curves never cross; smallest Stability gap %s points",
          fmt(min(g$gap)))
})

r_null <- res(!is.null(nfp), {
  b <- nfp[nfp$mode == "biomass", ]; o <- nfp[nfp$mode == "occurrence", ]
  sprintf("False-positive rate: C %s / %s%%, H′ %s / %s%%, Bs %s / %s%% (biomass / occurrence)",
          fmt(b$fpr_comp, 2), fmt(o$fpr_comp, 2), fmt(b$fpr_H, 2),
          fmt(o$fpr_H, 2), fmt(b$fpr_Bs, 2), fmt(o$fpr_Bs, 2))
})

r_cal <- res(!is.null(cvn) && !is.null(csv), {
  s <- cvn %>% filter(contrast == "PTa", family == "Substitution", mode == "biomass")
  g <- csv %>% filter(family == "Stability", mode == "biomass")
  sprintf("Substitution %s%% → %s%%; Stability gap %s → %s points (biomass)",
          fmt(s$pct_nominal), fmt(s$pct_calibrated),
          fmt(g$gap_nominal), fmt(g$gap_calibrated))
})

r_res <- res(!is.null(m5), {
  s <- m5 %>% filter(contrast == "PTa")
  sprintf("Median support %s (biomass) and %s (occurrence) across 100 resolutions",
          fmt(s$median_support[s$mode == "biomass"], 2),
          fmt(s$median_support[s$mode == "occurrence"], 2))
})

r_bal <- res(!is.null(bal), sprintf(
  "Bands disjoint in %s%% of draws; median gap %s–%s points",
  fmt(min(bal$pct_replicates_separated), 0),
  fmt(min(bal$median_gap_pp)), fmt(max(bal$median_gap_pp))))

r_eff <- res(nrow(eff) > 0, sprintf(
  "Gaps of %s–%s points under a common-support restriction, %d scale(s)",
  fmt(min(eff$gap_common)), fmt(max(eff$gap_common)), n_distinct(eff$level)))

tribble(
  ~Check,                          ~`Objection addressed`,                                ~Result,
  "Bootstrap over cells",          "Too few cells to separate the bands",                 r_boot,
  "Benjamini–Hochberg correction", "Uncorrected multiple testing",                        r_fdr,
  "Threshold sweep (α 0.01–0.075)","Dependence on α = 0.05",                              r_alpha,
  "Resolution stability",          "Dependence on prey aggregation",                      r_res,
  "Permutation null",              "Unequal calibration of the three tests",              r_null,
  "Calibrated thresholds",         "Classification driven by a liberal test",             r_cal,
  "Equalized sampling effort",     "Decadal contrast has more trawl sets",                r_bal,
  "Common-support restriction",    "Unequal effort, all three scales",                    r_eff
) %>% show_tbl()
```

**Table S8.** False-positive rates of the three tests and null frequencies of the diagnostic families, under within-cell permutation of period labels (200 permutations, between-period contrast, Gulf-wide scale).

```{r ts8}
nfp <- rd("NullCalibration_all_gulf_1/null_false_positive_rates.csv")
nfd <- rd("NullCalibration_all_gulf_1/null_family_distribution.csv")
if (is.null(nfp)) absent("NullCalibration_all_gulf_1/null_false_positive_rates.csv") else {
  nfp %>% relabel() %>%
    transmute(Currency = mode, `Null tests` = n,
              `Composition (%)` = fpr_comp, `Diversity H′ (%)` = fpr_H,
              `Niche breadth Bs (%)` = fpr_Bs, `Dispersion (%)` = fpr_disp) %>%
    show_tbl(digits = 2)
}
if (!is.null(nfd)) {
  cat("\n")
  nfd %>% relabel() %>% arrange(mode, family) %>%
    transmute(Currency = mode, Family = family,
              `Null mean (%)` = null_mean, `2.5%` = null_q025, `97.5%` = null_q975) %>%
    show_tbl()
}
```

**Table S9.** Significance thresholds calibrated on the permutation null: the fifth percentile of each test's null *p*-value distribution, overall and by class of sample size (the smaller of the two per-period set counts).

```{r ts9}
ca  <- rd("Calibrated_all_gulf_1/calibrated_alphas.csv")
cab <- rd("Calibrated_all_gulf_1/calibrated_alphas_by_sample_size.csv")
if (is.null(cab)) absent("Calibrated_all_gulf_1/calibrated_alphas_by_sample_size.csv") else {
  overall <- if (is.null(ca)) NULL else
    ca %>% transmute(mode, n_bin = "All", n = n_null, alpha_C, alpha_H, alpha_Bs)
  bind_rows(overall, cab) %>% relabel() %>%
    transmute(Currency = mode, `Sets per period` = n_bin, `Null tests` = n,
              `α composition` = alpha_C, `α diversity (H′)` = alpha_H,
              `α niche breadth (Bs)` = alpha_Bs) %>%
    show_tbl(digits = 3)
}
```

**Table S10.** Univariate effect sizes by diagnostic family, between-period contrast (2004–2006 vs 2018–2019), Gulf-wide scale. Medians and 90th percentiles across cells and prey resolutions.

```{r ts10}
m1 <- rd("Robustness_all_gulf_1_T/M1_effect_sizes_by_family.csv")
if (is.null(m1)) absent("Robustness_all_gulf_1_T/M1_effect_sizes_by_family.csv") else {
  m1 %>% filter(contrast == "PTa") %>% relabel() %>% arrange(mode, family) %>%
    transmute(Currency = mode, Family = family, `Cell × resolution` = n,
              `|ΔH′| median` = dH_med, `|ΔH′| 90th` = dH_q90,
              `|ΔBs| median` = dBs_med, `|ΔBs| 90th` = dBs_q90,
              `Bray–Curtis` = BC_med, `PERMANOVA R²` = R2_med) %>%
    show_tbl(digits = 3)
}
```

**Table S11.** Sampling effort per contrast and separation of the within- and between-period bands under a restriction to the effort range common to all five contrasts. Effort is the median number of trawl sets per period across cells.

```{r ts11}
eff_a <- map_dfr(names(LEVEL_LAB), function(l) {
  d <- rd(sprintf("Effort_%s_1/A_effort_by_contrast.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})
if (!nrow(eff_a)) absent("Effort_<level>_1/A_effort_by_contrast.csv") else {
  eff_a %>% filter(mode == "biomass") %>% relabel() %>%
    mutate(Scale = factor(LEVEL_LAB[level], levels = LEVEL_LAB),
           `Smaller group (IQR)` = sprintf("%s (%s–%s)", fmt(sets_min_med),
                                           fmt(sets_min_q25), fmt(sets_min_q75))) %>%
    arrange(Scale, contrast) %>%
    transmute(Scale, Contrast = contrast, Cells = n_cells,
              `Period 1` = sets_P1_med, `Period 2` = sets_P2_med,
              `Smaller group (IQR)`) %>%
    show_tbl()
}
eff_c <- map_dfr(names(LEVEL_LAB), function(l) {
  d <- rd(sprintf("Effort_%s_1/C_separation_verdict.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})
if (nrow(eff_c)) {
  cat("\n")
  eff_c %>% relabel() %>%
    mutate(Scale = factor(LEVEL_LAB[level], levels = LEVEL_LAB)) %>%
    arrange(Scale, mode, family) %>%
    transmute(Scale, Currency = mode, Family = family,
              `Gap, all cells` = gap_all, `Gap, common support` = gap_common) %>%
    show_tbl()
}
```

**Table S12.** Separation of the within- and between-period bands under equalized sampling effort. Proportion of independent draws in which the between-period band lies entirely outside the within-period band, and the gap between bands in percentage points.

```{r ts12}
t12 <- map_dfr(c("all_gulf", "ecoregion"), function(l) {
  d <- rd(sprintf("Robustness_%s_1_BAL/balanced_separation_verdict.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})
if (!nrow(t12)) absent("Robustness_<level>_1_BAL/balanced_separation_verdict.csv") else {
  t12 %>% relabel() %>%
    mutate(Scale = factor(LEVEL_LAB[level], levels = LEVEL_LAB)) %>%
    arrange(Scale, mode, family) %>%
    transmute(Scale, Currency = mode, Family = family, Draws = n_rep,
              `Draws separated (%)` = pct_replicates_separated,
              `Median gap` = median_gap_pp, `2.5% gap` = q025_gap_pp) %>%
    show_tbl()
}
```

**Table S13.** Frequency (%) of the four diagnostic families by contrast, currency and spatial scale: mean across the 100 prey resolutions, with the standard deviation across resolutions in parentheses. At the ecoregion and stratum scales, percentages are computed per spatial unit and averaged across units.

```{r ts13}
t13 <- map_dfr(names(LEVEL_LAB), function(l) {
  d <- rd(sprintf("Robustness_%s_1_T/M0_family_frequencies_reference.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})
if (!nrow(t13)) absent("Robustness_<level>_1_T/M0_family_frequencies_reference.csv") else {
  t13 %>% relabel() %>%
    mutate(Scale = factor(LEVEL_LAB[level], levels = LEVEL_LAB),
           cell = sprintf("%s (%s)", fmt(pct_mean), fmt(sd_across_x))) %>%
    select(Scale, Contrast = contrast, Currency = mode, family, cell) %>%
    pivot_wider(names_from = family, values_from = cell) %>%
    select(Scale, Contrast, Currency, any_of(FAMILIES)) %>%
    arrange(Scale, Currency, Contrast) %>%
    show_tbl()
}
```
