---
title: "A diagnostic framework for assessing trophic change from predator stomach contents"
subtitle: "Appendices"
format:
  docx:
    fig-width: 6.5
    fig-cap-location: bottom
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      results = "asis", dpi = 300, out.width = "100%")
source("R_helpers/paper_helpers.R")
```

# Appendix A. Prey taxonomic resolution

The primary analysis was repeated at 100 prey aggregation thresholds, from 10 to 1,000 stomachs in steps of 10 (Section 2.2; Figure 2). The tables below report the quantities cited in the Results.

**Table A1.** Compositional effect sizes by contrast and dietary currency, Gulf-wide scale: median Bray–Curtis dissimilarity between period-mean relative prey profiles and median PERMANOVA *R*², across classifiable cells and prey resolutions.

```{r ta1}
# Displayed as Table A1; the rebuilt file is TableA2_effect_sizes.csv.
show_pipeline("A2", digits = 3)
```

**Table A2.** Family frequencies at the ecoregion and stratum scales after excluding spatial units represented by fewer than five classifiable predator × size-class cells.

```{r ta2}
# Displayed as Table A2; the rebuilt file is TableA6_sparse_units.csv.
show_pipeline("A6")
```

**Table A3.** Range of Gulf-wide family frequencies across the 100 prey resolutions, by contrast and currency. The between-period range of Stability lies entirely below the lowest within-period value at any resolution.

```{r ta3}
# Displayed as Table A3; the rebuilt file is TableA5_resolution_range.csv.
show_pipeline("A5")
```

# Appendix B. Alternative prey-grouping rules

Two alternatives to the primary pooled rule were evaluated (Section 2.6): a cross-predator rule retaining only prey categories found in at least two predator species, and a per-predator rule building prey categories within each predator's diet. Comparisons use the 25 thresholds shared by the three rules.

**Table B1.** Gulf-wide family frequencies under the three grouping rules, by contrast and currency, on the 25 shared thresholds.

```{r tb1}
show_pipeline("B1")
```

**Table B2.** As Table B1, ecoregion scale.

```{r tb2}
show_pipeline("B2")
```

**Table B3.** As Table B1, stratum scale.

```{r tb3}
show_pipeline("B3")
```

```{r fb1, fig.cap="Figure B1. Gulf-wide family frequencies under the three grouping rules, by contrast and currency, on the 25 shared thresholds."}
fig_safe("^FigB1_")
```

```{r fb2, fig.cap="Figure B2. As Figure B1, ecoregion scale."}
fig_safe("^FigB2_")
```

```{r fb3, fig.cap="Figure B3. As Figure B1, stratum scale."}
fig_safe("^FigB3_")
```

# Appendix C. Robustness of the diagnostic assignment

This appendix reports the checks described in Section 2.6 and summarized in Section 3.4, all at the Gulf-wide scale under the primary grouping rule unless stated. Sampling uncertainty was estimated by resampling cells, multiple testing by Benjamini–Hochberg adjustment, test calibration by permuting period labels within cells, and sampling effort by equalizing and by restricting the number of trawl sets per period.

**Table C1.** Family frequencies (%) with 95% bootstrap intervals, obtained by resampling predator × size-class cells with replacement (2,000 replicates), by contrast and currency. Each entry gives the mean and the interval.

```{r tc1}
b <- rd("Robustness_all_gulf_1_T/M3_bootstrap_CI_family_pct.csv")
if (is.null(b)) absent("Robustness_all_gulf_1_T/M3_bootstrap_CI_family_pct.csv") else {
  b %>% relabel() %>%
    mutate(cell = sprintf("%s [%s–%s]", fmt(pct_mean), fmt(ci_lo), fmt(ci_hi))) %>%
    select(Contrast = contrast, Currency = mode, family, cell) %>%
    pivot_wider(names_from = family, values_from = cell) %>%
    select(Contrast, Currency, any_of(FAMILIES)) %>%
    arrange(Currency, Contrast) %>% show_tbl()
}
```

**Table C2.** Family frequencies (%) before and after Benjamini–Hochberg adjustment of the *p*-values of each test, applied within each combination of contrast, currency and resolution. Each entry gives the unadjusted and the adjusted value.

```{r tc2}
m2 <- rd("Robustness_all_gulf_1_T/M2_FDR_family_frequencies.csv")
if (is.null(m2)) absent("Robustness_all_gulf_1_T/M2_FDR_family_frequencies.csv") else {
  m2 %>% relabel() %>%
    mutate(cell = sprintf("%s → %s", fmt(pct_raw), fmt(pct_bh))) %>%
    select(Contrast = contrast, Currency = mode, family, cell) %>%
    pivot_wider(names_from = family, values_from = cell) %>%
    select(Contrast, Currency, any_of(FAMILIES)) %>%
    arrange(Currency, Contrast) %>% show_tbl()
}
```

```{r fc1, fig.cap="Figure C1. Calibration of the three tests under a permutation null. Period labels were permuted among trawl sets within each cell of the full between-period contrast, preserving the number of sets per period, so that no between-period difference exists by construction; the rejection rate at α = 0.05 is the empirical false-positive rate of each test. Rates are shown by class of sample size (the smaller of the two per-period set counts), from 200 permutations over ten prey resolutions. Dashed line: nominal 5%."}
fig_safe("^FigC1_")
```

**Table C3.** False-positive rates of the three tests and null frequencies of the diagnostic families under within-cell permutation of period labels (200 permutations, full between-period contrast).

```{r tc3}
nfp <- rd("NullCalibration_all_gulf_1/null_false_positive_rates.csv")
nfd <- rd("NullCalibration_all_gulf_1/null_family_distribution.csv")
if (is.null(nfp)) absent("NullCalibration_all_gulf_1/null_false_positive_rates.csv") else {
  nfp %>% relabel() %>%
    transmute(Currency = mode, `Null tests` = n,
              `Composition (%)` = fpr_comp, `Diversity H′ (%)` = fpr_H,
              `Niche breadth Bs (%)` = fpr_Bs, `Dispersion (%)` = fpr_disp) %>%
    show_tbl(digits = 2)
}
if (!is.null(nfd)) {
  nfd %>% relabel() %>% arrange(mode, family) %>%
    transmute(Currency = mode, Family = family, `Null mean (%)` = null_mean,
              `2.5th percentile` = null_q025, `97.5th percentile` = null_q975) %>%
    show_tbl()
}
```

```{r fc2, fig.cap="Figure C2. Frequency of the four diagnostic families under the nominal significance threshold (α = 0.05 for all three tests, grey) and under thresholds calibrated on the permutation null of Figure C1 (blue). Means across the 100 prey resolutions."}
fig_safe("^FigC2_")
```

**Table C4.** Significance thresholds calibrated on the permutation null: the fifth percentile of the null *p*-value distribution of each test, overall and by class of sample size (the smaller of the two per-period set counts).

```{r tc4}
ca  <- rd("Calibrated_all_gulf_1/calibrated_alphas.csv")
cab <- rd("Calibrated_all_gulf_1/calibrated_alphas_by_sample_size.csv")
if (is.null(cab)) absent("Calibrated_all_gulf_1/calibrated_alphas_by_sample_size.csv") else {
  overall <- if (is.null(ca)) NULL else
    ca %>% transmute(mode, n_bin = "All", n = n_null, alpha_C, alpha_H, alpha_Bs)
  bind_rows(overall, cab) %>% relabel() %>%
    transmute(Currency = mode, `Sets per period` = n_bin, `Null tests` = n,
              `α composition` = alpha_C, `α diversity (H′)` = alpha_H,
              `α niche breadth (Bs)` = alpha_Bs) %>%
    show_tbl(digits = 3)
}
```

```{r fc3, fig.cap="Figure C3. Family frequencies under equalized sampling effort, Gulf-wide scale. Each cell of each contrast was capped at the smallest per-period number of trawl sets observed among the within-period contrasts of that cell, sets were drawn without replacement, and the analysis was repeated over ten replicates. Boxes: distribution across replicates; crosses: values from the full data. Cells absent from any contrast or with a target below three sets were excluded (14 of the 65 cells in the set inventory)."}
fig_safe("^FigC3_")
```

**Table C5.** Separation of the within- and between-period frequencies under equalized sampling effort: proportion of replicates in which the between-period value lies entirely outside the within-period range, and the gap between them in percentage points.

```{r tc5}
t5 <- map_dfr(c("all_gulf", "ecoregion"), function(l) {
  d <- rd(sprintf("Robustness_%s_1_BAL/balanced_separation_verdict.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})
if (!nrow(t5)) absent("Robustness_<level>_1_BAL/balanced_separation_verdict.csv") else {
  t5 %>% relabel() %>%
    mutate(Scale = factor(LEVEL_LAB[level], levels = LEVEL_LAB)) %>%
    arrange(Scale, mode, family) %>%
    transmute(Scale, Currency = mode, Family = family, Replicates = n_rep,
              `Replicates separated (%)` = pct_replicates_separated,
              `Median gap` = median_gap_pp, `2.5th percentile gap` = q025_gap_pp) %>%
    show_tbl()
}
```

```{r fc4, fig.cap="Figure C4. As Figure C3, ecoregion scale. At this scale 132 of the 200 cells in the set inventory were excluded, most because they were absent from at least one contrast, and the median target was six sets per period; the balanced comparison therefore covers a limited subset of the assemblage."}
fig_safe("^FigC4_")
```

**Table C6.** Sampling effort per contrast and separation of the within- and between-period frequencies under a restriction to the effort range common to all five contrasts. Effort is the median number of trawl sets per period across cells; the gap is in percentage points.

```{r tc6}
eff_a <- map_dfr(names(LEVEL_LAB), function(l) {
  d <- rd(sprintf("Effort_%s_1/A_effort_by_contrast.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})
if (!nrow(eff_a)) absent("Effort_<level>_1/A_effort_by_contrast.csv") else {
  eff_a %>% filter(mode == "biomass") %>% relabel() %>%
    mutate(Scale = factor(LEVEL_LAB[level], levels = LEVEL_LAB),
           `Smaller group (IQR)` = sprintf("%s (%s–%s)", fmt(sets_min_med),
                                           fmt(sets_min_q25), fmt(sets_min_q75))) %>%
    arrange(Scale, contrast) %>%
    transmute(Scale, Contrast = contrast, Cells = n_cells,
              `Period 1` = sets_P1_med, `Period 2` = sets_P2_med,
              `Smaller group (IQR)`) %>%
    show_tbl()
}
eff_c <- map_dfr(names(LEVEL_LAB), function(l) {
  d <- rd(sprintf("Effort_%s_1/C_separation_verdict.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})
if (nrow(eff_c)) {
  eff_c %>% relabel() %>%
    mutate(Scale = factor(LEVEL_LAB[level], levels = LEVEL_LAB)) %>%
    arrange(Scale, mode, family) %>%
    transmute(Scale, Currency = mode, Family = family,
              `Gap, all cells` = gap_all, `Gap, common effort range` = gap_common) %>%
    show_tbl()
}
```

**Table C7.** Summary of the robustness checks: the question each addresses and its result. Values are read directly from the analysis outputs.

```{r tc7}
m3  <- rd("Robustness_all_gulf_1_T/M3_CI_overlap_verdict.csv")
m2  <- rd("Robustness_all_gulf_1_T/M2_FDR_family_frequencies.csv")
m4  <- rd("Robustness_all_gulf_1_T/M4_alpha_trend_sensitivity.csv")
nfp <- rd("NullCalibration_all_gulf_1/null_false_positive_rates.csv")
cvn <- rd("Calibrated_all_gulf_1/calibrated_vs_nominal.csv")
csv <- rd("Calibrated_all_gulf_1/calibrated_separation_verdict.csv")
bal <- rd("Robustness_all_gulf_1_BAL/balanced_separation_verdict.csv")
eff <- map_dfr(names(LEVEL_LAB), function(l) {
  d <- rd(sprintf("Effort_%s_1/C_separation_verdict.csv", l))
  if (is.null(d)) NULL else mutate(d, level = l)
})
res <- function(ok, expr) if (isTRUE(ok)) expr else "Not run"

r_boot <- res(!is.null(m3), sprintf(
  "Stability intervals separated by %s (biomass) and %s (occurrence) points",
  fmt(m3$gap_pp[m3$mode == "biomass" & m3$family == "Stability"]),
  fmt(m3$gap_pp[m3$mode == "occurrence" & m3$family == "Stability"])))
r_fdr <- res(!is.null(m2), {
  s <- m2 %>% filter(contrast == "PTa", family == "Substitution", mode == "biomass")
  sprintf("Between-period Substitution %s%% → %s%% (biomass)", fmt(s$pct_raw), fmt(s$pct_bh))
})
r_alpha <- res(!is.null(m4), {
  g <- m4 %>% filter(trend == 0.10, alpha <= 0.075, family == "Stability") %>%
    group_by(mode, alpha, contrast_type) %>%
    summarise(lo = min(pct_mean), hi = max(pct_mean), .groups = "drop") %>%
    pivot_wider(names_from = contrast_type, values_from = c(lo, hi)) %>%
    mutate(gap = lo_within - hi_between)
  sprintf("No crossing for α = 0.01–0.075; smallest Stability gap %s points", fmt(min(g$gap)))
})
r_null <- res(!is.null(nfp), {
  b <- nfp[nfp$mode == "biomass", ]; o <- nfp[nfp$mode == "occurrence", ]
  sprintf("False-positive rates: C %s / %s%%, H′ %s / %s%%, Bs %s / %s%% (biomass / occurrence)",
          fmt(b$fpr_comp, 2), fmt(o$fpr_comp, 2), fmt(b$fpr_H, 2), fmt(o$fpr_H, 2),
          fmt(b$fpr_Bs, 2), fmt(o$fpr_Bs, 2))
})
r_cal <- res(!is.null(cvn) && !is.null(csv), {
  s <- cvn %>% filter(contrast == "PTa", family == "Substitution", mode == "biomass")
  g <- csv %>% filter(family == "Stability", mode == "biomass")
  sprintf("Substitution %s%% → %s%%; Stability gap %s → %s points (biomass)",
          fmt(s$pct_nominal), fmt(s$pct_calibrated), fmt(g$gap_nominal), fmt(g$gap_calibrated))
})
r_bal <- res(!is.null(bal), sprintf(
  "Separated in %s%% of replicates; median gap %s–%s points",
  fmt(min(bal$pct_replicates_separated), 0),
  fmt(min(bal$median_gap_pp)), fmt(max(bal$median_gap_pp))))
r_eff <- res(nrow(eff) > 0, sprintf(
  "Gaps of %s–%s points at %d scale(s)",
  fmt(min(eff$gap_common)), fmt(max(eff$gap_common)), n_distinct(eff$level)))

tribble(
  ~Check,                          ~`Question addressed`,                               ~Result,
  "Bootstrap over cells",          "Is the separation larger than sampling error?",     r_boot,
  "Benjamini–Hochberg adjustment", "Does repeated testing drive the families?",         r_fdr,
  "Threshold sweep",               "Does the result depend on α = 0.05?",               r_alpha,
  "Permutation null",              "Are the three tests equally calibrated?",           r_null,
  "Calibrated thresholds",         "Does a liberal test drive the classification?",     r_cal,
  "Equalized sampling effort",     "Does the decadal contrast gain from more sets?",    r_bal,
  "Common effort range",           "Does unequal effort drive the result at all scales?", r_eff
) %>% show_tbl()
```
