---
title: "A diagnostic framework for assessing trophic change from predator stomach contents"
subtitle: "Tables"
format:
  docx: default
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      results = "asis")
source("R_helpers/paper_helpers.R")
```

**Table 1.** Diagnostic typology. Each predator × size-class cell is scored on three tests — prey composition (C, PERMANOVA), diversity (*H*′, jackknifed Shannon) and niche breadth (*B*s, standardized Levins) — as significant (sig, *p* < 0.05) or non-significant (ns). A composition trend (0.05 ≤ *p* < 0.10) is used only when *H*′ and *B*s are both non-significant; otherwise the cell is classified on the significant signal.

```{r t1}
tribble(
  ~`Diagnostic state`,           ~C,              ~`H′`, ~Bs,   ~Family,
  "Stable Diet",                 "ns (p ≥ 0.10)", "ns",  "ns",  "Stability",
  "Emerging Shift",              "trend",         "ns",  "ns",  "Stability",
  "Ghost Shift",                 "sig",           "ns",  "ns",  "Substitution",
  "Internal Rebalancing",        "ns",            "sig", "ns",  "Functional change",
  "Niche Compression/Expansion", "ns",            "ns",  "sig", "Functional change",
  "Niche Restructuring",         "ns",            "sig", "sig", "Functional change",
  "Partial Diet Shift",          "sig",           "sig", "ns",  "Reorganization",
  "Structural Shift",            "sig",           "ns",  "sig", "Reorganization",
  "Major Shift",                 "sig",           "sig", "sig", "Reorganization"
) %>% show_tbl()
```

**Table 2.** Gulf-wide frequency (%) of the four diagnostic families by contrast and dietary currency. Values are means across the 100 prey taxonomic resolutions of the primary grouping rule, with the standard deviation across resolutions in parentheses, computed over classifiable cells.

```{r t2}
d <- freq_wide("all_gulf")
if (is.null(d)) absent("Robustness_all_gulf_1_T/M0_family_frequencies_reference.csv") else show_tbl(d)
```

**Table 3.** Frequency (%) of the four diagnostic families at the ecoregion and stratum scales, by contrast and dietary currency. Frequencies are computed within each spatial unit, with a family absent from a unit counted as zero, then averaged across units and across the 100 prey taxonomic resolutions; standard deviation across resolutions in parentheses.

```{r t3}
t3 <- map_dfr(c("ecoregion", "stratum"), function(l) {
  d <- freq_wide(l)
  if (is.null(d)) NULL else mutate(d, Scale = LEVEL_LAB[[l]], .before = 1)
})
if (!nrow(t3)) absent("Robustness_<ecoregion|stratum>_1_T/M0_family_frequencies_reference.csv") else show_tbl(t3)
```

# Supplementary tables

**Table S1.** Predator taxa retained in the analysis, with the length at maturity (*L*~mat~) used to separate juvenile and adult size classes (Scharf et al., 2000, and regional sources). Taxa without a published regional *L*~mat~ were analysed as a single size class.

```{r ts1}
show_pipeline("S1")
```

**Table S2.** Predator × size-class cells classifiable in the full between-period contrast (2004–2006 vs 2018–2019), with the numbers of non-empty stomachs and trawl sets in each sampling window. Two further cells, juvenile snakeblenny and adult thorny skate, were classifiable only in within-period contrasts.

```{r ts2}
show_pipeline("S2")
```

**Table S3.** Frequency (%) of the four diagnostic families within each ecoregion, by contrast and dietary currency, with the number of classifiable predator × size-class cells in each ecoregion.

```{r ts3}
# Displayed as Table S3; the pipeline file is still named TableS4_*.
show_pipeline("S4")
```

**Table S4.** Frequency (%) of the four diagnostic families within each survey stratum, by contrast and dietary currency, with the number of classifiable predator × size-class cells in each stratum.

```{r ts4}
# Displayed as Table S4; the pipeline file is still named TableS5_*.
show_pipeline("S5")
```
